A New Unitary RG-based Auxiliary Model Approach to Strongly-Correlated Systems

### 2022-04-28 gen-siam-manuscript/notes.tex
Added k-space self-energy to notes

This expression can be seen as the sum of effects of creating a single-particle excitation \(c^\dagger_\sigma\) on the ground state \(\ket{\Psi_\text{gs} }\), at increasingly larger distances \(\vec R_j\) from the auxiliary model at \(\vec R_0\). The excitation is made to evolve through the auxiliary model \(\mathcal{H}_\text{aux}\left( \vec R_0 \right) \) centred at \(\vec R_0\). The excitation is ultimately annihilated and the state returns to the ground state. However, since we are describing the excitation of the bulk ground state using an auxiliary model Hamiltonian, we also need to multply this with the fidelity between the full ground state and the auxiliary model ground state. That is taken care of by the inner-product in the numerator multiplying the matrix element. That inner product can be thought of as an overlap between a separable state and an entangled state - the full Hamiltonian state is obtained by entangling the overlapping auxiliary models, while the single auxiliary model ground state is the separable state in this context. The inner product in the numerator then becomes a measure of entanglement of the bulk ground state - the more the overlap with the separable state, the lower the entanglement: \(\varepsilon(\vec R_0,\Psi_\text{gs}) \equiv 1 - |\braket{\psi^0_\text{aux}\left(\vec R_0\right) | \Psi_\text{gs}}|^2 \). The inner product in the denominator is also an entanglement: \(\varepsilon(\vec R_0, \ket{\vec k}) = 1 - |\braket{\psi^0_\text{aux}\left(\vec R_0\right) | \ket{\vec k}}|^2\). This can be thought of as the entanglement within each auxiliary model ground state, with the momentum eigenstates \(\ket{\vec k}\) acting as the separable states.
	\braket{i | H_\text{H-H} | i} = N \left( 1 - \varepsilon(\vec R_0,\vec k) \right) \bra{\Phi_0(\vec k)} \sum_{\vec R_j} c_{\sigma}\left(\vec R_j\right)\mathcal{H}_\text{aux}\left( \vec R_0 \right)c^\dagger_{\sigma}\left(\vec R_j\right)\ket{\Phi_0(\vec k)}
The more is the entanglement content within the groundstate, the lower is the contribution from a single auxiliary model \(\mathcal{H}_\text{aux}\left( \vec R_0 \right)\). This makes sense between the contributions will then be distributed across all the auxiliary models. On the other hand, when the entanglement is low, any single auxiliary model provides a large contribution, which can be understood from the fact that any single auxiliary model is sufficient to describe the full ground state.

@@

\section{Important Remarks}
\subsection{Entanglement as an order parameter for the transition}
The spectral weights in the diagonal Greens function can be written in terms of a geometric measure of entanglement:
\begin{equation}\begin{aligned}
	d_n^p = \braket{\Phi_n | c^\dagger_{d\sigma} | \Phi_0} = \braket{\Phi_n | c^\dagger_{d\sigma} \sum_n \ket{\phi_m} \bra{\phi_m} | \Phi_0}
\end{aligned}\end{equation}
where, \(\ket{\phi_m}\) are the eigenstates of the Hamiltonian composed of a two-site system (formed by the impurity site and the zeroth site) in direct product with the rest of the lattice sites. The two site system interacts only via a spin-exchange coupling \(J\), and the impurity and bath sites have local correlations \(U\) and \(U_b\) respectively. The ground state is the spin-singlet in direct product with the ground state of the tight-binding chain formed by the remaining sites. The summation involves inner products of the ground state \(\ket{\Phi_0}\) with the eigenstates \(\ket{\phi_m}\). Only two states out of all \(\ket{\phi_m}\) give non-zero overlap - the spin singlet state \(\ket{SS}\) and the charge triplet-zero state \(\ket{CT}\). These two non-zero overlaps give rise to non-zero entanglement: \(\varepsilon_\text{SS} = 1 - |\braket{\Phi_0 | SS}|^2, \varepsilon_\text{CT} = 1 - |\braket{\Phi_0 | CT}|^2\). The spectral weight becomes
\begin{equation}\begin{aligned}
	d_n^p = \bra{\Phi_n} | c^\dagger_{d\sigma} \left(\ket{SS} + \ket{CT} \right) \left(\sqrt{1 - \varepsilon_\text{SS}} + \sqrt{1 - \varepsilon_\text{CT}}\right)
\end{aligned}\end{equation}
The total spectral weight can be written as
\begin{equation}\begin{aligned}
	|d_n^p|^2 = |\bra{\Phi_n} | c^\dagger_{d\sigma} \left(\ket{SS} + \ket{CT} \right)|^2 \left(2 - \varepsilon_\text{SS} - \varepsilon_\text{CT} + 2\sqrt{1 - \varepsilon_\text{SS}}\sqrt{1 - \varepsilon_\text{CT}}\right)
\end{aligned}\end{equation}
Near the metal, the ground state is, to a very good approximation, the state \(\ket{SS}\). One can then say that the entanglement \(\varepsilon_\text{SS}\) is poor - the entangled state is very close to the "separable" state. The term separable is used here in the sense that the metallic state defines a limit of the system. As \(U_b\) increases, the e

>>>>

This expression can be seen as the sum of effects of creating a single-particle excitation \(c^\dagger_\sigma\) on the ground state \(\ket{\Psi_\text{gs} }\), at increasingly larger distances \(\vec R_j\) from the auxiliary model at \(\vec R_0\). The excitation is made to evolve through the auxiliary model \(\mathcal{H}_\text{aux}\left( \vec R_0 \right) \) centred at \(\vec R_0\). The excitation is ultimately annihilated and the state returns to the ground state. However, since we are describing the excitation of the bulk ground state using an auxiliary model Hamiltonian, we also need to multply this with the fidelity between the full ground state and the auxiliary model ground state. That is taken care of by the inner-product in the numerator multiplying the matrix element. That inner product can be thought of as an overlap between a separable state and an entangled state - the full Hamiltonian state is obtained by entangling the overlapping auxiliary models, while the single auxiliary model ground state is the separable state in this context. The inner product in the numerator then becomes a measure of entanglement of the bulk ground state - the more the overlap with the separable state, the lower the entanglement: \(\varepsilon(\ket{\Psi_\text{gs}}) \equiv 1 - |\braket{\psi^0_\text{aux}\left(\vec R_0\right) | \Psi_\text{gs}}|^2 \). The inner product in the denominator is also an entanglement: \(\varepsilon( \ket{\vec k}) = 1 - |\braket{\psi^0_\text{aux}\left(\vec R_0\right) | \ket{\vec k}}|^2\). This can be thought of as the entanglement within each auxiliary model ground state, with the momentum eigenstates \(\ket{\vec k}\) acting as the separable states.
	\braket{i | H_\text{H-H} | i} = \frac{1 - \varepsilon\left( \ket{\Psi_\text{gs}}\right)}{1 - \varepsilon\left( \vec k\right)}\sum_{\vec R_j} \bra{\Psi_\text{gs}} c_{\vec R_j,\sigma}\mathcal{H}_\text{aux}\left( \vec R_0 \right)c^\dagger_{\vec R_j,\sigma}\ket{\Psi_\text{gs}}
The more is the entanglement content within the groundstate, the lower is the contribution from a single auxiliary model \(\mathcal{H}_\text{aux}\left( \vec R_0 \right)\). This makes sense between the contributions will then be distributed across all the auxiliary models. On the other hand, when the entanglement is low, any single auxiliary model provides a large contribution, which can be understood from the fact that any single auxiliary model is sufficient to describe the full ground state. The numerator, therefore, is maximum in the metal and minimum (zero) in the insulator. Similarly, the denominator is minimum in the metal. This is because, in the metal, the auxiliary model ground states are delocalised, so the entanglement between multiple \(\vec k\)-states will be lower compared to the insulating phase, where the ground state is completely localised. This shows the entire fraction in front is maximum in the metal and minimum (zero) in the insulator.

@@

\section{\(k-\)space picture of the MIT in terms of the momentum-resolved self-energy \(\Sigma(\vec k,\omega)\)}

From the self-energy expressions obtained above, one can track the self-energy of the lattice model through the metal-insulator transition. In fig.~\ref{four-self}, we have provided these plots at specific values of \(U_b/J\) to show how the transition plays out in the Brillouin zone. {\it These plots can also be viewed as a video \href{https://abhirup-m.github.io/misc-sigma/}{\underline{here}}.}

\begin{center}
\begin{tabular}{c|c}
\includegraphics[height=0.3\textheight]{../figures/sigma-4-Ub_by_J=-0.02500-4.pdf} &
\includegraphics[height=0.3\textheight]{../figures/sigma-4-Ub_by_J=-0.10310-4.pdf}\\
\hline
\includegraphics[height=0.3\textheight]{../figures/sigma-4-Ub_by_J=-0.15735-4.pdf} &
\includegraphics[height=0.3\textheight]{../figures/sigma-4-Ub_by_J=-0.23594-4.pdf}\\
\hline
\includegraphics[height=0.3\textheight]{../figures/sigma-4-Ub_by_J=-0.25000-4.pdf} &
\includegraphics[height=0.3\textheight]{../figures/sigma-4-Ub_by_J=0.25010-4.pdf}
\end{tabular}
\captionof{figure}{Each of the 6 frames shows \(A(\vec k)\) at four values of \(\omega\). First 4 are before the transition (\(-U_b/J < 0.25\)), fifth is exactly at the transition \((-U_b/J = 0.25)\) and last one is after the transition (\(-U_b/J > 0.25\)). Available as video \href{https://abhirup-m.github.io/misc-sigma/}{\underline{here}}.}
\label{four-self}
\end{center}

In each frame, the top left figure represents the self-energy at \(\omega=0\). If we track this particular figure across the frames, we find that at small \(-U_b/J\), the self-energy near the Fermi surface was smaller than those near the Brillouin zone edges, indicating the presence of a prominent Fermi liquid. As \(U_b\) increases, we find that this distinction of the Fermi surface becomes less-prominent; at the critical point,we see that the self-energy near the Fermi surface self-energy has become more than that near the edges, signalling that the quasiparticle excitations have become scrambled because of the increased correlations in the bath.

If we, on the other hand, track the bottom right figure in each frame (\(\omega = U/2\)), we find that deep in the metallic regime, the self-energy is quite high. This is because of the low spectral weight at large \(\omega\). As the correlations are increased, we find that  this self-energy decreases, particular in comparison to the values at the Brillouin zone edges. This is because of the appearance of the Hubbard side-peaks in the local spectral function
.

In the insulating phase, we find that the momentum-dependence has vanished (the system is in the perfectly local atomic limit). The self-energy at the Hubbard side-bands (bottom right) is high, because the quasiparticle excitations are highly-renormalised and the single-particle picture description is not possible. The self-energy inside the gap (other three figures) diverges and the poles of the Greens function have been replaced by zeroes, so we have instead plotted the inverse self-energy.

These plots are available as videos \href{https://abhirup-m.github.io/misc-sigma/}{\underline{here}}.

\clearpage
\section{Some important remarks}
\subsection{Entanglement as an order parameter for the transition}
The spectral weights in the diagonal Greens function can be written in terms of a geometric measure of entanglement:
\begin{equation}\begin{aligned}
	d_n^p = \braket{\Phi_n | c^\dagger_{d\sigma} | \Phi_0} = \braket{\Phi_n | c^\dagger_{d\sigma} \sum_n \ket{\phi_m} \bra{\phi_m} | \Phi_0}
\end{aligned}\end{equation}
where, \(\ket{\phi_m}\) are the eigenstates of the Hamiltonian composed of a two-site system (formed by the impurity site and the zeroth site) in direct product with the rest of the lattice sites. The two site system interacts only via a spin-exchange coupling \(J\), and the impurity and bath sites have local correlations \(U\) and \(U_b\) respectively. The ground state is the spin-singlet in direct product with the ground state of the tight-binding chain formed by the remaining sites. The summation involves inner products of the ground state \(\ket{\Phi_0}\) with the eigenstates \(\ket{\phi_m}\). Only two states out of all \(\ket{\phi_m}\) give non-zero overlap - the spin singlet state \(\ket{SS}\) and the charge triplet-zero state \(\ket{CT}\). These two non-zero overlaps give rise to non-zero entanglement: \(\varepsilon_\text{SS} = 1 - |\braket{\Phi_0 | SS}|^2, \varepsilon_\text{CT} = 1 - |\braket{\Phi_0 | CT}|^2\). The spectral weight becomes
\begin{equation}\begin{aligned}
	d_n^p = \bra{\Phi_n} | c^\dagger_{d\sigma} \left(\ket{SS} + \ket{CT} \right) \left(\sqrt{1 - \varepsilon_\text{SS}} + \sqrt{1 - \varepsilon_\text{CT}}\right)
\end{aligned}\end{equation}
The total spectral weight can be written as
\begin{equation}\begin{aligned}
	|d_n^p|^2 = |\bra{\Phi_n} | c^\dagger_{d\sigma} \left(\ket{SS} + \ket{CT} \right)|^2 \left(2 - \varepsilon_\text{SS} - \varepsilon_\text{CT} + 2\sqrt{1 - \varepsilon_\text{SS}}\sqrt{1 - \varepsilon_\text{CT}}\right)
\end{aligned}\end{equation}
Near the metal, the ground state is, to a very good approximation, the state \(\ket{SS}\). One can then say that the entanglement \(\varepsilon_\text{SS}\) is poor - the entangled state is very close to the "separable" state. The term separable is used here in the sense that the metallic state defines a limit of the system. As \(U_b\) increases, \(\varepsilon_\text{SS}\) increases. On the other hand, the entanglement with the charge triplet is maximum in the metal, because the ground state does not have much of the charge triplet. Towards the MIT, this contribution increases, and the entanglement decreases.
